Biden Expected to Win 54% of Popular Vote in 2020 US Presidential Election

### 2020-11-02 mrp.Rmd

```{r}
race_prop <- cell_counts %>% 
  ungroup %>%
  group_by(race_ethnicity) %>%
  mutate(prop = n/sum(n)) %>%
  ungroup()

gender_prop <- cell_counts %>% 
  ungroup %>%
  group_by(gender) %>%
  mutate(prop = n/sum(n)) %>%
  ungroup()

state_prop <- cell_counts %>% 
  ungroup %>%
  group_by(state) %>%
  mutate(prop = n/sum(n)) %>%
  ungroup()

age_prop <- cell_counts %>% 
  ungroup %>%
  group_by(age_group) %>%
  mutate(prop = n/sum(n)) %>%
  ungroup()
```

```{r}
race_res <- readRDS("processed_data/race_res.RDS")
gender_res <- readRDS("processed_data/gender_res.RDS")
state_res <- readRDS("processed_data/state_res.RDS")
age_res <- readRDS("processed_data/age_res.RDS")

@@

```{r}
age_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(age_group), color = "MRP estimate")) + 
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) + 
  ylab("proportion keeping name") + 
  xlab("age in 2020") + 
  geom_point(data = training_data %>% 
               group_by(age_group, vote_biden) %>%
               dplyr::summarise(n = n()) %>% 
               group_by(age_group) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(age_group, prop, color = "MNCS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "MNCS raw data" = "red")) + 
  theme_bw(base_size = 14) 
ggsave("images/age_plot.png", width = 8, height = 6)
```
```{r}
race_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(race_ethnicity), color = "MRP estimate")) + 
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) + 
  ylab("proportion keeping name") + 
  xlab("age in 2020") + 
  geom_point(data = training_data %>% 
               group_by(race_ethnicity, vote_biden) %>%
               dplyr::summarise(n = n()) %>% 
               group_by(race_ethnicity) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(race_ethnicity, prop, color = "MNCS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "MNCS raw data" = "red")) + 
  theme_bw(base_size = 14) 
ggsave("images/race_plot.png", width = 8, height = 6)
```
```{r}
state_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(state), color = "MRP estimate")) + 
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) + 
  ylab("proportion keeping name") + 
  xlab("age in 2020") + 
  geom_point(data = training_data %>% 
               group_by(state, vote_biden) %>%
               dplyr::summarise(n = n()) %>% 
               group_by(state) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(state, prop, color = "MNCS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "MNCS raw data" = "red")) + 
  theme_bw(base_size = 14) 
ggsave("images/race_plot.png", width = 8, height = 6)
```

>>>>

```{r}
#load the summarized results for each variable we used
race_res <- readRDS("processed_data/race_res.RDS")
gender_res <- readRDS("processed_data/gender_res.RDS")
state_res <- readRDS("processed_data/state_res.RDS")
age_res <- readRDS("processed_data/age_res.RDS")
education_rest <- readRDS("processed_data/age_res.RDS")

@@

```{r}
age_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(age_group), color = "MRP estimate")) + 
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) + 
  ylab("proportion keeping name") + 
  xlab("age in 2020") + 
  geom_point(data = training_data %>% 
               group_by(age_group, vote_biden) %>%
               dplyr::summarise(n = n()) %>% 
               group_by(age_group) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(age_group, prop, color = "MNCS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "MNCS raw data" = "red")) + 
  theme_bw(base_size = 14) 
```

```{r}
race_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(race_ethnicity), color = "MRP estimate")) + 
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) + 
  ylab("proportion keeping name") + 
  xlab("age in 2020") + 
  geom_point(data = training_data %>% 
               group_by(race_ethnicity, vote_biden) %>%
               dplyr::summarise(n = n()) %>% 
               group_by(race_ethnicity) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(race_ethnicity, prop, color = "MNCS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "MNCS raw data" = "red")) + 
  theme_bw(base_size = 14) 
```

```{r}

state_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(state), color = "MRP estimate")) + 
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) + 
  ylab("proportion keeping name") + 
  xlab("age in 2020") + 
  geom_point(data = training_data %>% 
               group_by(state, vote_biden) %>%
               dplyr::summarise(n = n()) %>% 
               group_by(state) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(state, prop, color = "MNCS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "MNCS raw data" = "red")) + 
  theme_bw() +
  coord_flip()

```{r}
#show selected states for more clarity
#include key, swing and heavy trump/heavy biden states for discussion

filtered_state_res <- state_res %>% filter(state == "PA" | state == "VT" | 
                                           state == "MI" | state == "AR" |
                                          state == "WI")
filtered_training_data <- training_data %>% filter(state == "PA" | state == "VT" | 
                                           state == "MI" | state == "AR" |
                                          state == "WI")
filtered_state_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(state), color = "MRP estimate")) + 
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) + 
  ylab("proportion keeping name") + 
  xlab("age in 2020") + 
  geom_point(data = filtered_training_data %>% 
               group_by(state, vote_biden) %>%
               dplyr::summarise(n = n()) %>% 
               group_by(state) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(state, prop, color = "MNCS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "MNCS raw data" = "red")) + 
  theme_bw() +
  coord_flip()
```

```{r}
gender_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(gender), color = "MRP estimate")) + 
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) + 
  ylab("proportion keeping name") + 
  xlab("Gender in 2020") + 
  geom_point(data = training_data %>% 
               group_by(gender, vote_biden) %>%
               dplyr::summarise(n = n()) %>% 
               group_by(gender) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(gender, prop, color = "MNCS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "MNCS raw data" = "red")) + 
  theme_bw(base_size = 14) 
```

```{r}
education_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(education), color = "MRP estimate")) + 
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) + 
  ylab("proportion keeping name") + 
  xlab("Gender in 2020") + 
  geom_point(data = training_data %>% 
               group_by(education, vote_biden) %>%
               dplyr::summarise(n = n()) %>% 
               group_by(education) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(education, prop, color = "MNCS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "MNCS raw data" = "red")) + 
  theme_bw(base_size = 14) 
```

### 2020-11-02 outputs/paper/paper.Rmd

>>>>

Since our predictor y is a binary variable (Biden or Trump), we used a logistic regression in order build our model. A logistic regression predicts the likelihood of an outcome based on various independent variables and it fits a model on a log curve which is bound in the Y axis in the range of [0,1].. In our case, we judge the likelihood that an individual would vote for Joe Biden in comparison to Donald Trump, given the individuals age, gender, education, state, and race. Given these inputs and outputs, we can train our logistic regression model using BRMS to predict which values for each variable predicts certain votes, and this will be discussed further in the later section. In our model, as it is a binary predictor, we round the prediction to the nearest number (more than 0.5 is rounded to 1, meaning this individual is likely to vote for Joe Biden, and less than 0.5 is rounded to 0, meaning this individual is likely to vote to Donald Trump).

One variable we discussed about including was annual income, however when we added the income lowered our cross validation score and the well discussed issues with self-reported income (bias, inaccurate estimations, not reporting) convinced us to disclude income. Another variable we debated on was education. Originally, education was included, however, education severely impacted our cross validation accuracy. We solved these issues by summarizing all education levels into 2 levels: at least some Post-Secondary and No post-secondary. This allowed our model to perform slightly better, but more importantly, there was a significant difference in the voting preference between the 2 groups. Another variable that helped increase accuracy after manipulation was age, where it became a better independent variable after breaking age into groups on 10's (eg. 18-28, 29-38) in order to better represent different age groups. Other groupings we tried was breaking into youth, middle age, and seniors (18-35, 36-55, 55+) and leaving age as individual numeric values, however both of these breakdowns did not provide any significant value.

Overall, our model performs fairly (Figure N), with a cross-validation accuracy of approximately 63% (Figure N). This performed better than 50% (guessing) and through this we are able to gain some insight on how each variable affect the prediction. Furthermore, by investigating our inaccuracies (Figure N), we see that of the approximately 21% of our predictions were false negatives (Predicted Trump vote when it should predict Biden) and 15% of our predictions were false positives (predicted Biden when it should have predicted Trump vote). This was significantly better than some of the other models we tried as discussed in the previous paragraphs, as we had a large number of false negatives in many models when education was split into multiple groups rather than just College or no College.

```{r}
model_education <- readRDS("../../model/4chains_3000iter_.rds")

reduced_data <- readRDS("../../inputs/training_data.Rda")

total_rows <- nrow(reduced_data)

# get updated columns (used for the model)
reduced_data <- reduced_data %>% select(vote_biden,age_group, education, gender,
                                        race_ethnicity, state)

set.seed(50)
#set
shuffle_indices <- sample(total_rows)
data <- reduced_data[shuffle_indices,]
boundary <- as.integer(total_rows*0.95)

training <- data[0:boundary,]

testing <- data[boundary:total_rows,]

#testing the accuracy of the model
probability <- predict(model_education, type = "response", newdata = testing)
probability <- if_else(probability[,1] >0.5,1,0)
testing$probs <- probability
testing <- testing %>% mutate(acc = probs==vote_biden)
data.frame(table(testing$acc)) %>% 
  ggplot(aes(x = Var1, y = Freq/(sum(Freq)))) +
  geom_bar(stat="identity") +
  xlab("Correct Prediction") +
  ylab("Percent") +
  ggtitle("Cross Validation Scores")
```

```{r}
wrong <- as.numeric(testing$vote_biden) - as.numeric(testing$probs) - 1

wrong <- plyr::mapvalues(wrong, c(-1,0,1), c("False Positive", 
                                                   "Correct",
                                                   "False Negative"))

data.frame(table(wrong)) %>% 
  ggplot(aes(x = wrong, y = Freq/sum(Freq)))+
  geom_bar(stat="identity") +
  xlab("Prediction Correctness") +
  ylab("Total")+
  ggtitle("Cross Validation inaccuracies")
